Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-08 kappa.tex
Add references.

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
\cref{fig:map-northamerica}. The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
the Western Cordillera, which forms our region of focus in the present thesis.

@@

In North American, the Western Cordillera forms a broad mountain system with an
average east-west width of about 1000\,km. Its topography is a complex
arrangement of mountain ranges and interior basins. It was formed during
multiple accretion phases associated with the closure of several back-arc
oceanic trenches over a 200\,Ma-long geological history \mref.

Most of the dominant reliefs are found along two major strips that run parallel
to each other along the general orientation of the Cordillera
\cref{fig:map-cordillera}. The western strip consists of the Pacific Coast
orogenic belt. Major ice-covered reliefs include,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern strip consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench. It includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
Brooks range, in northern Alaska, has been covered by ice, but its glaciers
were not connected to the main body of the Cordilleran ice sheet.

The Cordillera comports several intermontane basins that impacted patterns
of ice sheet flow. The two major depressions are the Interior Plateau of
central British Columbia, and the Liard Lowland. Between these two basins is
found a mildly elevated, but extensive and deeply incised mountain group, known
as the Skeena Mountains.

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
of the irregular arrangement of valleys and ridges that formed the ice sheet's
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them \mref. In
terms of
numerical glacier modelling, it implies complex patterns of flow and stress
within the ice, and requires high-resolution, and thus, large computational
grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effect is often
clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
large latitudinal extent of the region, mean annual temperatures vary
significantly from north to south (\cref{fig:plot-atm}a). About half of the
model domain has mean annual temperature below freezing. However, seasonality,
here computed as the temperature difference between the warmest and the coldest
months, varies between around 10\,\degC\ along most of the Pacific Coast to
above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exist a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperature is well below
freezing point do experience warm summers.

Precipitation rates are highest on the western slopes of the Pacific Coast
ranges and decline quickly on the other side of the mountains
(\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
even reinforced by the difference in timing of the precipitation peak through
the year (\cref{fig:plot-atm}d). Coastal regions experience most precipitation
during winter and fall, thus as snow. On the contrary, many inland regions
experience dry winters, and most of the annual precipitation consequently falls
as rain during the ablation season. The climatic setting of the North American
Cordillera is described further in \CCLI, where its effects on the ice sheet
model response have been studied in detail.

@@

During the last glacial maximum, the Cordilleran ice sheet covered most of the
Canadian Cordillera, and some of the adjacent land in north-western contiguous
USA and Alaska (\cref{map-cordillera}). The southern margin is the most
extensively studied. It consists of different piedmont lobes that have formed
at the outlet of the main valleys. The Puget lobe was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \mref. As the ice sheet retreated, the lake repeatedly
and catastrophically drained to the Pacific Ocean \mref. These floods
contributed to shape a distinctive landscape known as the Channeled Scablands.

>>>>

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
(\cref{fig:map-northamerica}). The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
the Western Cordillera, which forms our region of focus in the present thesis.

@@

The North American Cordillera forms a broad mountain system with an
average east-west width of about 1000\,km. Its topography is a complex
jigsaw of mountain ranges and interior basins (\cref{fig:map-cordillera}).
It was formed during multiple accretion phases associated with the closure of
several back-arc oceanic trenches over a 200\,Ma-long geological history
\citep{Sigloch.Mihalynuk.2013}.

Most of the dominant reliefs are found along two major ribbons that run
parallel to each other along the general orientation of the Cordillera
(\cref{fig:map-cordillera}). The western ribbon is formed by the Pacific Coast
orogenic belt. Major glaciated reliefs include,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench \mref. It includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
Brooks Range, in northern Alaska, has been covered by ice, but its glaciers
were not connected to the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

\todo{Go to geolibrary and check \emph{Geology of Canada} books and such for a
      reference about the three orogenic belts.}

The Cordillera comports several intermontane basins, which influenced the
patterns of ice flow. The two major depressions are the Fraser Plateau in
south-central British Columbia, and the Liard Lowland in northern British
Columbia. Between these two basins is found a mildly elevated, but extensive
and deeply incised mountain group, known as the Skeena Mountains.

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
of the irregular arrangement of valleys and ridges that formed the ice sheet's
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them
\citep{Davis.Mathews.1944, Kleman.etal.2010}. In terms of numerical glacier
modelling, this topography implies complex patterns of flow and stress
within the ice, and requires high-resolution, and thus, large computational
grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effect is often
clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
large latitudinal extent of the region, mean annual temperatures vary
significantly from north to south (\cref{fig:plot-atm}a). About half of the
model domain has mean annual temperature below freezing. However, seasonality,
here computed as the temperature difference between the warmest and the coldest
months, varies between around 10\,\degC\ along most of the Pacific Coast to
above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperature is well below
freezing point do experience warm summers.

Precipitation rates are highest on the western slopes of the Pacific Coast
ranges and decline quickly on the other side of the mountains
(\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
even reinforced by the difference in timing of the precipitation peak through
the year (\cref{fig:plot-atm}d). Coastal regions receive most precipitation
during winter and fall, thus as snow. On the contrary, many inland regions
experience dry winters, and most of the annual precipitation, there,
consequently falls
as rain during the ablation season. The climatic setting of the North American
Cordillera is described further in \CCLI, where its effects on the ice sheet
model response have been studied in detail.

@@

During the last glacial maximum, the Cordilleran ice sheet covered most of the
Canadian Cordillera, and some of the adjacent land in north-western contiguous
USA and Alaska (\cref{fig:map-cordillera}). The southern margin is the most
extensively studied. It consists of different piedmont lobes that have formed
at the outlet of the main valleys. The Puget lobe was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape known as the Channeled Scabland.
